Whole-chromosome pairwise alignment using the hierarchical wavefront algorithm

### 2021-06-14 manuscript/main.tex
added why not seed-and-extend mapping methods

%\title[short Title]{Wavefront inception scales global pairwise alignment to whole eukaryotic chromosomes}
%\title[short Title]{Wavefront inception enables global pairwise alignment to whole eukaryotic chromosomes}
    \title[wfmash: whole-cromosome pairwise alignment]{Whole-chromosome pairwise alignment using a hierarchical wavefront algorithm}
    \author[Sample \textit{et~al}.]{
        Corresponding Author\,$^{\text{\sfb 1,}*}$,
        Co-Author\,$^{\text{\sfb 2}}$ and
        Co-Author\,$^{\text{\sfb 3,}*}$
    }
    \address{
        $^{\text{\sf 1}}$Department, Institution, City, Post Code, Country \\
        $^{\text{\sf 2}}$Department, Institution, City, Post Code, Country \\
        $^{\text{\sf 3}}$Department, Institution, City, Post Code, Country
    }

@@

    \abstract{
        \textbf{Motivation:} Pairwise alignment of sequences is an important step in many bioinformatics analyses,
        including pangenomes building. Pangenomes are sequence models able to provide a full representation of the
        mutual alignment of collections of genomes. The time and memory required to compute a pairwise alignment
        increase quadratically with the sequence length, making it impractical to directly align very long sequences
        without applying heuristic approaches to determine possible syntenic regions in the sequences. Nevertheless,
        with the advances in sequencing technologies, new genome assemblies are produced at a high rate, pressing for
        the development of tools able to align sequences of the order of tens of megabases long.
        \\
        \textbf{Results:} Here we present \wfmash\/, a new gap-affine pairwise aligner designed to align DNA sequences
        at a whole-chromosome scale. \wfmash\ applies a hierarchical implementation of the wavefront alignment
        algorithm to guide the alignment of very long sequences. It scales efficiently to large eukaryotic
        chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little
        time and memory.
        \\
        \textbf{Availability:} \wfmash\ is written in \Cpp\ programming language. Source code and user manual are freely available at
        \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
        \\
        \textbf{Contact:}
        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}
        \\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
    }

@@

    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping
    \citep{22388286, BWA_MEM, 23103880} to variant detection~\citep{21478889}, as well as de~novo
    assembly \citep{19251739} and pangenome building~\citep{33177663, 33066802}. Pairwise sequence alignment
    can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary
    relationships between two biological sequences. In this context, pangenomes model the full set
    of genomic elements in a given species or clade~\citep{32453966}. These data structures encode the mutual
    relationships between all the genomes represented, in contrast to reference-based approaches which
    relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
    the entire genetic diversity of a population. FIXME In combination with the massive amount
    of data available made possible thanks to the advancement in sequencing technology, presses for the
    development of tools able to compute pairwise alignment of very long sequences.

    To face the upcoming challenges in pangenomes building, we have developed \wfmash\/, a new gap-affine
    pairwise tool for aligning DNA sequences at a whole-chromosome scale. It supports split-read alignment
    and gap-affine penalties for insertion and deletions. \wfmash\ scales efficiently to large eukaryotic
    chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise
    alignment of hundreds of genomes. \wfmash\ is a DNA sequence aligner based on the integration of
    MashMap~\citep{30423094}, a fast approximate aligner for computing local alignment boundaries between
    long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952}, an exact gap-affine algorithm
    that takes advantage of homologous regions between the sequences to accelerate the alignment process.

    %TODO Some sentences of why it exists with respect to Minimap2/Winnomap2, explaining the limitations of those tools in the pairwise alignment task.

@@

    To avoid such limitation, wfmash applies a hierarchical WFA, exploiting the WFA to guide the alignment
    process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
    sequences, it aligns them to each other in small pieces \textit{W}-bps long by applying a global WFA alignment
    (Figure~1\vphantom{\ref{fig:1}}). The whole global alignment is then computed over the full
    dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
    Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments
    \textit{W}-bps long from the two sequences to align. To determine if each cell is a match, and then guiding the
    \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the
    two fragments.

@@

    We evaluated the performance of \wfmash\ 0.xx against Minimap2-v2.20-r1061 \citep{29750242} and Winnomap 2.03 \citep{XXX}, in
    terms of runtime and memory usage. The comparison was performed on both simulated and real datasets. Simulated
    data allowed evaluating the alignment quality by calculating precision and recall in variant identification.
    The evaluations were performed on a workstation running Ubuntu 20.04 equipped with 64 GB of RAM and an Intel i9-9900K
    CPU with 8 cores (16 threads). All commands are provided in the Supplementary Material.

    \subsection{Simulated data}
    We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and
    the telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of divergence and length
    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool~\citep{splitfa},
    In each run, all sequences have the same length, with 50\% of them in reverse complement.
    Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced using the
    Mutation-Simulator script~\citep{32780803}.
    %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.

@@

    We implemented a novel gap-affine pairwise aligner, \wfmash\/, to accelerate the computation of the mutual
    alignment of collections of genomes, a step required for building pangenome models. We have demonstrated that
    it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the
    Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, wfmash is already
    successfully applied in our pangenome graphs building pipeline (\citep{pggb}). No less important, pairwise
    alignment is a central step of many bioinformatics applications, making our aligner a scalable solution to face the
    increasing yields of sequencing technologies in the coming years.

>>>>

%\title[short Title]{Wavefront inception scales global pairwise alignment to whole eukaryotic chromosomes}
%\title[short Title]{Wavefront inception enables global pairwise alignment to whole eukaryotic chromosomes}
    \title[wfmash: whole-cromosome pairwise aligner]{Whole-chromosome pairwise alignment using the hierarchical wavefront algorithm}
    \author[Sample \textit{et~al}.]{
        Corresponding Author\,$^{\text{\sfb 1,}*}$,
        Co-Author\,$^{\text{\sfb 2}}$ and
        Co-Author\,$^{\text{\sfb 3,}*}$
    }
    \address{
        $^{\text{\sf 1}}$Department, Institution, City, Post Code, Country \\
        $^{\text{\sf 2}}$Department, Institution, City, Post Code, Country \\
        $^{\text{\sf 3}}$Department, Institution, City, Post Code, Country
    }

@@

    \abstract{
        \textbf{Motivation:} Pairwise alignment of sequences is an important step in many bioinformatics analyses,
        including pangenome construction. Pangenomes are sequence models able to provide a full representation of the
        mutual alignment of collections of genomes. The time and memory required to compute pairwise alignments
        increase quadratically with the sequence length, making it impractical to directly align very long sequences
        without applying heuristic approaches to first determine possible syntenic regions in the sequences. Nevertheless,
        with the advances in sequencing technologies, new genome assemblies are produced at a high rate, pressing for
        the development of tools able to align sequences of the order of tens of megabases long.
        \\
        \textbf{Results:} Here we present \wfmash\/, a new gap-affine pairwise aligner designed to align DNA sequences
        at a whole-chromosome scale. \wfmash\ applies a hierarchical implementation of the wavefront alignment
        algorithm to guide the alignment of very long sequences. It scales efficiently to large eukaryotic
        chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little
        time and memory.
        \\
        \textbf{Availability:} \wfmash\ is written in \Cpp\ programming language. Source code and user manual are freely
        available at \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
        \\
        \textbf{Contact:}
        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}
        \\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
    }

@@

    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{22388286, BWA_MEM, 23103880}
    to variant detection~\citep{21478889}, as well as de~novo assembly~\citep{19251739} and
    pangenome construction~\citep{33177663, 33066802}. Pairwise sequence alignment can be applied to identify
    similar regions that may indicate functional, structural, and/or evolutionary
    relationships between two biological sequences. In this context, pangenomes model the full set
    of genomic elements in a given species or clade~\citep{32453966}. These data structures encode the mutual
    relationships between all the genomes represented, in contrast to reference-based approaches which
    relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
    the entire genetic diversity of a population. This opportunity, in combination with the massive amount of data
    available nowadays, presses for the development of tools able to compute pairwise alignment of very long sequences.

    Popular seed-and-extend mapping methods, like Minimap2~\citep{29750242}, poorly scale in runtime and memory
    when generating sensitive alignments for chromosome-scale contigs available thanks to the new sequencing methods
    like PacBio HiFi. To scale to vertebrate genomes, methods for pangenome construction based on these approaches
    must first filter out centromeric and other highly-repetitive sequences.

    In the interest of operating on the full pangenome, and to face the upcoming challenges in pangenome construction,
    we have developed \wfmash\/, a new gap-affine pairwise tool for aligning DNA sequences at a whole-chromosome scale.
    It supports split-read alignment and gap-affine penalties for insertion and deletions. \wfmash\ scales
    efficiently to large eukaryotic chromosomes while requiring little time and memory, making it suitable to
    be applied on the pairwise alignment of hundreds of genomes. \wfmash\ is a DNA sequence aligner based on the
    integration of MashMap~\citep{30423094}, a fast approximate sequence mapper for computing local alignment boundaries
    between long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952}, an exact gap-affine
    algorithm that takes advantage of homologous regions between the sequences to accelerate the alignment process.

@@

    To avoid such limitation, \wfmash\ applies a hierarchical WFA, exploiting the WFA to guide the alignment
    process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
    sequences, it aligns them to each other in small pieces \textit{W}-bps long by applying a global WFA alignment
    (Figure~1\vphantom{\ref{fig:1}}). The whole global alignment is then computed over the full
    dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
    Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments
    \textit{W}-bps long from the two sequences to align. To determine if each cell is a match, and then guiding the
    \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the
    two fragments.

@@

    We evaluated the performance of \wfmash\ \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{29750242} and Winnomap 2.03~\citep{XXX}, in
    terms of runtime and memory usage. The comparison was performed on both simulated and real datasets. Simulated
    data allowed evaluating the alignment quality by calculating precision and recall in variant identification.
    The evaluations were performed on a workstation running Ubuntu 20.04 equipped with 64 GB of RAM and an Intel i9-9900K
    CPU with 8 cores (16 threads). All commands are provided in the Supplementary Material.

    \subsection{Simulated data}
    We simulated long sequences from the telomere-to-telomere (T2T) assemblies of the human chromosomes 8,
    at different levels of length and divergence (Table 1 and Table 2). Chromosomes were split into shorter
    sequences using the splitfa tool~\citep{splitfa}, In each run, all sequences have the same length, with 50\%
    of them in reverse complement. Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs)
    were introduced using the Mutation-Simulator script~\citep{32780803}.
    %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.

@@

    We implemented a novel gap-affine pairwise aligner, \wfmash\/, to accelerate the computation of the mutual
    alignment of collections of genomes, a step required for constructing pangenome models. We have demonstrated that
    it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the
    Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, \wfmash\ is already
    successfully applied in our pangenome graphs building pipeline (\citep{pggb}). No less important, pairwise
    alignment is a central step of many bioinformatics applications, making our aligner a scalable solution to face the
    increasing yields of sequencing technologies in the coming years.
